---
title: "Intro to data visualisation"
subtitle: ""
author: "Jakub Nowosad"
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, scipen = 99*99)
```

## Data visualisation in R

- R has several systems for making graphics - **base R**, the **lattice** package, and the **ggplot2** package are among the most popular ones

---
## **ggplot2** - basic information

- **ggplot2** is an implementation of the grammar of graphics (["Grammar of Graphics"](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448) book written by Leland Wilkinson)
- data frames are an input data (most of the times)
- the **ggplot2** package has two functions for plot creation - `qplot()` and `ggplot()`
- **ggplot2** documentation is available at http://docs.ggplot2.org

```{r, eval=FALSE}
install.packages("ggplot2")
```

```{r}
library('ggplot2')
```

---
## **ggplot2** - the first plot

```{r, fig.align='center', fig.height=6}
df <- data.frame(A = c(1:10), B = seq(22, 4, by= -2))
ggplot(data = df, aes(x = A, y = B)) + geom_point()
```

---
## ggplot2 - basic dictionary

- `geom` - type of a plot ("histogram", "boxplot", "point", etc.)
- aesthetics (`aes`) - visual properties of geoms, such as color, size, shape, ...
- faceting (`facet`) - panels with a different subset of the data 

---
## Dataset

```{r, eval=FALSE}
# install.packages('devtools')
# devtools::install_github("jennybc/gapminder")
install.packages('gapminder')
library(gapminder)
```

or

```{r}
gapminder <- readRDS('data/gapminder.rds')
```

http://www.gapminder.org/data/

http://github.com/jennybc/gapminder

http://www.youtube.com/watch?v=jbkSRLYSojo

```{r}
gapminder2007 <- subset(gapminder, year==2007)
```

---
## Histogram

```{r, tidy=FALSE, message=FALSE, fig.align='center', fig.height=6}
ggplot(data=gapminder2007, aes(x=gdpPercap)) + 
        geom_histogram()
```

---
## Histogram 

.pull-left[
```{r, tidy=FALSE, message=FALSE, warning=FALSE, out.width =  '400px'}
ggplot(data=gapminder2007, aes(x=gdpPercap)) + 
        geom_histogram(binwidth=10000)
```
]

.pull-right[
```{r, tidy=FALSE, message=FALSE, warning=FALSE, out.width =  '400px'}
ggplot(data=gapminder2007, aes(x=gdpPercap)) + 
        geom_histogram(binwidth=800)
```
]

---
## Bar plot

- There are two types of bar plots in **ggplot2**:
    - `geom_bar()` makes the height of the bar proportional to the number of cases in each group
    - `geom_col()` makes the height of the bars to represent values in the data

```{r, fig.align='center',fig.height=4}
ggplot(data=gapminder2007, aes(x=continent)) + geom_bar()
```

---
## Bar plot

- There are two types of bar plots in **ggplot2**:
    - `geom_bar()` makes the height of the bar proportional to the number of cases in each group
    - `geom_col()` makes the height of the bars to represent values in the data

```{r, fig.align='center',fig.height=4, message=FALSE, warning=FALSE}
library('dplyr')
gapminder2007_2 <- gapminder2007 %>% group_by(continent) %>% summarise(number=n())
ggplot(data=gapminder2007_2, aes(x=continent, y=number)) + geom_col()
```

---
## Line plot
```{r, fig.align='center',fig.height=6}
gapminder2 <- gapminder %>% group_by(year) %>% summarise(mean.lifeExp=mean(lifeExp, na.rm=TRUE))
ggplot(data=gapminder2, aes(x=year, y=mean.lifeExp)) + geom_line()
```

---
## Scatter plot

```{r, fig.align='center',fig.height=6}
ggplot(data=gapminder2007, aes(x=gdpPercap, y=lifeExp)) + geom_point()
```

---
## Box plot

```{r, fig.align='center',fig.height=6}
ggplot(data=gapminder2007, aes(x=continent, y=lifeExp)) +
        geom_boxplot()
```

---
## Aesthetic attributes for quantitative variables (1)

```{r, fig.align='center',fig.height=6}
ggplot(data=gapminder2007, aes(x=gdpPercap, y=lifeExp, color=pop)) +
        geom_point()
```

---
## Aesthetic attributes for quantitative variables (2)

```{r, fig.align='center',fig.height=6}
ggplot(data=gapminder2007, aes(x=gdpPercap, y=lifeExp, size=pop)) +
        geom_point()
```

---
## Aesthetic attributes for qualitative variables (1)

```{r, fig.align='center',fig.height=6}
ggplot(data=gapminder2007, aes(x=gdpPercap, y=lifeExp, color=continent)) +
        geom_point()
```

---
## Aesthetic attributes for qualitative variables (2)

```{r, fig.align='center',fig.height=6}
ggplot(data=gapminder2007, aes(x=gdpPercap, y=lifeExp, shape=continent)) + 
        geom_point() + scale_shape(solid = FALSE)
```

---
## Scales 

```{r, fig.align='center', fig.height=6}
ggplot(data=gapminder2007, aes(x=gdpPercap, y=lifeExp, color=pop)) +
        geom_point()+ 
        scale_colour_gradientn(colours=rainbow(5))
```

---
## Data transformations

```{r, fig.align='center', fig.height=6}
ggplot(data=gapminder2007, aes(x=gdpPercap, y=lifeExp, color=log10(pop))) +
        geom_point()
```

---
## Facets

- Two functions, `facet_grid()` and `facet_wrap()` can be used to plot the subsets of data together

```{r, fig.align='center',fig.height=5, message=FALSE}
ggplot(data=gapminder2007, aes(x=lifeExp)) + 
        geom_histogram() + 
        facet_wrap(~continent)
```

---
## Plot customizing - labs

```{r, tidy=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.height=5}
p <- ggplot(data=gapminder2007, aes(x=lifeExp, fill=continent)) +
        geom_histogram()
p <- p + labs(x='Life expectancy', y=NULL,
              title='Histogram', fill='Continent: ')
p
```

---
## Plot customizing - themes

```{r, tidy=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.height=4}
p <- p + theme_bw()
p
```

---
## Saving plots

```{r, tidy=FALSE, message=FALSE, eval=FALSE}
p <- ggplot(data=gapminder2007, aes(x=lifeExp, fill=continent)) +
        geom_histogram() +
        labs(x='Life expectancy', y=NULL,
              title='Histogram', fill='Continent: ') +
        theme_bw()
```

```{r, tidy=FALSE, message=FALSE, eval=FALSE}
ggsave(filename = "Plot.pdf", plot = p)
```

```{r, tidy=FALSE, message=FALSE, eval=FALSE}
ggsave(filename = "Plot.png", plot = p, dpi = 300)
```

---
## That's not all (folks)

- More geoms:
   - `geom_density()`
   - `geom_jitter()`
   - `geom_text()`
   - ... 
- Axes and legends customizing
- Additional packages (**ggplot2** extensions) - http://www.ggplot2-exts.org/
- Spatial data visualisation
- ... and many more

---
## What's next?

- Interactive plots (plotly, dygraph, ...)

<!--https://github.com/bhaskarvk/widgetframe-->

--- 

Resources:

- http://varianceexplained.org/RData/lessons/lesson2/
- http://www.njtierney.com/Resources/
- http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/
- http://www.cookbook-r.com/Graphs/
- http://shiny.stat.ubc.ca/r-graph-catalog/
- https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf